A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-25 code/supplements.Rmd
update figure references to reflect new kNN clusters, update figure 3 results, add APOBEC to supplement

##oncogene
oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
##tumor_supressor
tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL") ## list of selected tumor suppressor

>>>>

##oncogene
oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
##tumor_supressor
tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL", "APOBEC3B") ## list of selected tumor suppressor

### 2020-07-25 text/manuscript.Rmd
updated figure2 to include figure 3. removed figure 3 code from manuscript
update figure references to reflect new kNN clusters, update figure 3 results, add APOBEC to supplement

It does not matter for our purposes whether a single tumor is “enriched” for
multiple pathway mutations, chiefly because such metrics favor larger pathways (
_i.e._ pathways with more member genes) over smaller ones, and non-expressed
genes over expressed ones due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. Data selected in
this manner are likely to be noisy for several reasons. First, point mutations
can be deleterious (attenuating or hypomorphic) or activating (neomorphic) in
genes that act functionally as either activators or repressors. Within the scope
of this study we must accept this uncertainty and assume mutations are
generically disruptive to the cell regardless of whether they are activating or
deactivating or whether they affect a gene that we might classify as one thing
or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, as previously
noted, we know that low-expressed genes and non-expressed genes are known to
accumulate mutations at a higher rate due to transcription coupled repair. To
address this issue, we identified genes with low expression in each type of
cancer and eliminated them for that cancer type only (see methods for details).
<strike>Highly expressed genes also may have high mutation rates owing to
transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
that this mechanism results in cell-type-specific biases that might be
biologically meaningful for predisposition to different classes of cancer
however, and therefore chose to include these genes in our analysis.</strike>
After selecting our pathways and genes, we then compiled a matrix of the
pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
member gene carries a mutation) to the pathway regardless of the number of genes
carrying mutations (this data is visualized in **figure 1a**).

```{r child='../code/figure1.Rmd'}
```
In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
containing useful information (**figure 1b**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several large, robust projections
reproducible regardless of random seed setting. A representative version of this
UMAP is shown in **figure 1c** as a 2D projection. Following this spatial
mapping we attempted to define groupings of similar tumors within the spatial
map using HDBSCAN, which performs hierarchical clustering and provides metrics
of cluster stability and probabilities of cluster membership for each node.
However, HDBSCAN is sensitive to several parameters; key for our analysis are
the minimum number of tumor samples in a cluster that capture the maximum number
of tumor samples, measured by probability of membership of $\geq 5\%$ in at
least one class. Thus, we created a score metric as the fraction of classified
tumors with max probability $< 5\%$ in one class and chose a cluster size of 95
to minimize the score function (**figure 1d**). HDBSCAN with these settings
resulted in 12 distinct high-density clusters which we then projected onto the
UMAP (**figure 1e**). This classified <mark>x out of 7,607 tumors</mark> but
still resulted in a significant fraction of unclassified tumor samples. Since we
ultimately wish to be able to classify any tumor using this scheme, we performed
k Nearest Neighbors (kNN) analysis, which computes a similarity metric to every
tumor in the set and then lets the $k$ most similar tumors "vote" as to the
identity of the query tumor sample based on their classification labels. We set
$k$ to be the square root of the number of tumor samples (87). Using this
method, we assigned cluster membership to the remaining tumors (**figure 1f**).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1h**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 9 & 10, which is
distinguished  by KRAS mutations among others. But even PAAD comprises tumors
from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. For example we projected annotations for each of the breast
cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
most heavily studied molecular subtypes in cancer, which each have different
prognoses and standards of care. We did not observe any exclusive segregation by
pathway for these subtype annotations in GDC (**Figure 1i** and **supplemental
file 1**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** for the full set of projections); we find that
the subtypes, though often biased towards one or more classes, are almost never
exclusive to one of the classes. We interpret these data in aggregate to mean
that our pathway disruption classes do not correspond to molecular subtypes
within the parent cancer type. Furthermore, this result raises the possibility
that, molecular and histo-pathological subtypes could represent different
precursor cell origins within each cancer type. This interpretation still leaves
open the possibility that different precursor cell types, being susceptible to
different teratogens and selection pressures ( _e.g._ insulin dependence) could
be biased toward one or more of the pathway disruption profiles.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability, proof-reading and DNA damage repair, and
telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
breast, prostate and ovarian cancers (refs). The mechanism of risk is thought to
involve loss of heterozygosity [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 1j**). We made
similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including diameter,
which can vary greatly in importance between different tissues. Stage is used
clinically as a proxy for advancement toward a more deadly state and metastasis.
Given these assumptions, it is possible that more advanced tumors have common
pathway disruption profiles, and the UMAP which features a series of lobe-like
structures on a common backbone of tumor samples (**figure 1c**) could in
principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone culminates in a group of tumors (class 8) which
has nearly every pathway disrupted. However, outside of class 8 we don't observe
an obvious trend in the mutation count across the backbone of the UMAP (**figure
1g**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
classes represent advancement through stages, we projected these data onto the
UMAP. Similar to tissue of origin and other categories of tumor, we did not
observe any bias amongst the stages into specific pathway disruption classes
(**figure 1k**), suggesting that stage is not an determining factor in pathway
disruption classes.

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met. These phenotypes include (among others)
loss of differentiation, cell-cell contacts, epithelial to mesenchyme
transition, immune system evasion and tissue invasiveness. To determine whether
any of our classes correspond to an especially advanced stage of cancer accross
tissue types, we projected the metastasis data onto our UMAP, and surprisingly,
again, we observed an even distribution across classes (**figure 1l**). This
final observation suggests that our pathway-disruption classification is
dependent on particular combinations of gene mutations that hit different
pathway profiles that can each give rise to advanced stages of disease and
metastasis, regardless and independent of overall mutational burden.

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
replacing genes with pathways (**figure 2**). A complete list of pathway
disruptions with percent mutated samples and top genes is provided in
**supplementary table ST1**). These reveal the broad importance of many well
known pathways that are disrupted in cancer, including <mark>PI3K/Akt
activation, DNA repair etc fix this with top 5 or so quoted pathway
labels</mark>. The top third of pathways drop off gradually in overall mutation
frequency, possibly owing to relative diversity among all tumors.

@@

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
0 pathways in class 8 (for which most pathways are putatively disrupted), 25
pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
had enrichment. To explore the specific pathways marking each class, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
metabolic pathways including RNA and protein biosynthesis (**supplementary
figure S4**). Most class 4 enriched pathways were not unique to this class but
shared in class 10 & 12, suggesting that this grouping reflects either greater
heterogeneity or possibly a lower stage of tumor evolution (as distinct from
clinical ``stage''). Classes 9-12 have in common mutations in extracellular
signaling, intracellular signaling, and immune-related pathways (see **figures
3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
distinct of the 12, we chose to investigate these two groups further.
<div style="text-align: right"> Figure 3 data </div>

```{r child='../code/figure3.Rmd'}
```

@@

<mark>This section will have to be rewritten to reflect Simon's new clusters.</mark>

Since pathways like these share many common genes, we investigated whether
tumors in this class  have multiple mutations across signaling pathways, or
whether a small number of driver genes account for enrichment. The split between
clusters 9 & 10 and 11 & 12 is largely driven by mutations in _PI3K_ and its
homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
**supplementary figure S4**). Furthermore, clusters 9 and 11 are distinguished
from 10 and 12 by general lack of P53 mutations (compare **Figure 1f** with the
_TP53_ panel of **Supplementary Figure S4**).

In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
the highest pathway enrichment scores of the three major categories, though
there is a significant number of tumors spatially distributed between cluster
groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
pre-replicative complex," "metabolism of polyamines," "degradation of
$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
with aspects of hedgehog signaling. To determine once again whether the genes
driving pathway enrichment were commonly shared among these pathways we plotted
and performed unsupervised clustering on tumors with an individual mutation
matrix using every gene in each of the ten pathways, grouped by shared pathway
membership (**figure 4**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity
contributes to all the aforementioned pathways and the majority of tumors in
these classes have a single mutation in one of these genes. Nonetheless, a
significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
proteasomal mutations but instead harbor mutations in genes representing a
majority of the same pathways, suggesting that most if not all of the pathways
are individually important and that proteasomal mutations may be sufficient to
drive the net effect of the other mutations in this cluster. We annotated the
tumors in this class according to tissue of origin to determine whether
proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
indeed we could observe some clustering. These data could explain how
tissue-specific gene mutation signatures might result in general cancer
phenotypes by converging at the pathway level.

@@

The proteasome plays key roles in the transduction of signals from the p53, WNT
and NFkB pathways and it and proteasomal inhibitors have been approved for
clinical use in some patient populations [@proteasome_review_Jang]. To determine
whether this class of proteins is broadly sensitive to cancer-promoting
mutations, we tabulated each gene as prognostic or not in the human protein
expression atlas [@protein_atlas] and found 49 positive and 10 negative
associations ( _i.e._ higher expression associated with unfavorable outcomes is
"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
Second we looked at survival in mutations results results results (**figure 4c**). 
Third, we checked the mutation classes and found the overwhelming
majority (x out of y) are missense _vs._ various other deletarious classes (see
**figure 4d**), suggesting that, similar to other oncogenic driver classes such
as RAS and PI3K, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.

>>>>

It does not matter for our purposes whether a single tumor is “enriched” for
mutations of individual pathways, chiefly because such metrics favor larger
pathways ( _i.e._ pathways with more member genes) over smaller ones. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic) in genes that act functionally as either activators or repressors.
Within the scope of this study we accept this uncertainty and assume mutations
are generically disruptive to the pathway regardless of whether they are
activating or deactivating or whether they affect a gene that we might classify
as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
know that low-expressed genes and non-expressed genes are known to accumulate
mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details).
<strike>Highly expressed genes could also have high mutation rates owing to
transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
that this mechanism results in cell-type-specific biases that might be
biologically meaningful for predisposition to different classes of cancer
however, and therefore chose not to exclude these genes from our analysis. After
selecting our pathways and genes, we then compiled a matrix of the pathways,
assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
carries a mutation) to the pathway regardless of the number of genes carrying
mutations (**figure 1A**).

```{r child='../code/figure1.Rmd'}
```
In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
containing useful information (**figures 1B, C**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several large, robust projections
reproducible regardless of random seed setting. A representative version of this
3D UMAP is shown in **figure 1D** as a 2D projection, rotated to enhance the
visibility of the major features. Following this spatial mapping we attempted to
define groupings of similar tumors within the spatial map using HDBSCAN, which
performs hierarchical clustering and provides metrics of cluster stability and
probabilities of cluster membership for each node. However, HDBSCAN is sensitive
to several parameters; key for our analysis are the minimum number of tumor
samples in a cluster that capture the maximum number of tumor samples, measured
by probability of membership of $\geq 5\%$ in at least one class. Thus, we
created a score metric as the fraction of classified tumors with max probability
$< 5\%$ in one class and chose a cluster size of 95 to minimize the score
function (**figure 1E**). HDBSCAN with these settings resulted in 12 distinct
high-density clusters which we then projected onto the UMAP (**figure 1F**).
This classified <mark>x out of 7,607 tumors</mark> but still resulted in a
significant fraction of unclassified tumor samples. Since we ultimately wish to
be able to classify any tumor using this scheme, we performed k Nearest
Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
the set and then lets the $k$ most similar tumors "vote" as to the identity of
the query tumor sample based on their classification labels. We set $k$ to be
the square root of the number of tumor samples (87). Using this method, we
assigned cluster membership to the remaining tumors (**figure 1G**).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1I**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 8 & 9, which is
distinguished  by KRAS mutations among others (**supplementary figure S3**). But even PAAD comprises tumors
from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. For example we projected annotations for each of the breast
cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
most heavily studied molecular subtypes in cancer, which each have different
prognoses and standards of care. We did not observe any exclusive segregation by
pathway for these subtype annotations (**Figure 1J** and **supplemental
file 1**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** for the full set of projections); we find that
the subtypes, though often biased towards one or more classes, are almost never
exclusive to one of the classes. We interpret these data in aggregate to mean
that our pathway disruption classes do not correspond to molecular subtypes
within the parent cancer type. Furthermore, this result raises the possibility
that, molecular and histo-pathological subtypes could represent different
precursor cell origins within each cancer type. This interpretation still leaves
open the possibility that different precursor cell types, being susceptible to
different teratogens and selection pressures ( _e.g._ insulin dependence) could
be biased toward one or more of the pathway disruption profiles.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability, proof-reading and DNA damage repair, and
telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
breast, prostate and ovarian cancers. The mechanism of risk is thought to
involve loss of heterozygosity [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 1K**). We made
similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including diameter,
which can vary greatly in importance between different tissues. Stage is used
clinically as a proxy for advancement toward a more deadly state and metastasis.
Given these assumptions, it is possible that more advanced tumors have common
pathway disruption profiles, and the UMAP which features a series of lobe-like
structures on a common backbone of tumor samples (**figure 1D**) could in
principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has the fewest point mutations and culminates in a cluster (class 10) which
has nearly every pathway disrupted (**figure 1H**). However, outside of class 10 we don't observe
an obvious trend in the overall mutation burden across the backbone of the UMAP (**figure
1H**). Nonetheless, to test the hypothesis that the molecular-pathway disruption
classes represent advancement through stages, we projected staging data onto the
UMAP. Similar to tissue of origin and other categories of tumor, we did not
observe any bias amongst the stages into specific pathway disruption classes
(**figure 1L**), suggesting that stage is not an determining factor in pathway
disruption classes.

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met. These phenotypes include (among others)
loss of differentiation, cell-cell contacts, epithelial to mesenchyme
transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To determine whether
any of our classes correspond to an especially advanced stage of cancer accross
tissue types, we projected the metastasis data onto our UMAP, and surprisingly we observed an even distribution across classes (**figure 1M**). This
final observation suggests that our pathway-disruption classification is
dependent on particular combinations of gene mutations that hit different
pathway profiles that can each give rise to advanced stages of disease and
metastasis, regardless and independent of overall mutational burden (see Discussion for further analysis).

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created waterfall graphs analagous to the oncoprint
visualizations on cBioportal [@cbioportal], replacing genes with pathways
(**figure 2A**). A complete list of pathway disruptions with percent mutated
samples and top genes is provided in **supplementary table ST1**). As expected,
these reveal the broad importance of many well known pathways that are disrupted
in cancer, including "PIP3 activates Akt signaling" (77% of samples),
"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
samples), _etc._ . The top third of pathways drop off gradually in overall
mutation frequency (**figure 1A** and **supplementary table ST1**), possibly
owing to relative diversity among all tumors.

@@

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 2B**). Using this approach, we
identified 25 pathways highly enriched (enrichment score > 0.3, see methods) in
class 2, 9 pathways enriched in class 3, 24 pathways in class 4, 10 pathways in
class 5, 10 pathways in class 6, 24 pathways in class 7, 24 pathways in class 8
(for which most pathways are putatively disrupted), 10 pathways in class 9, and
0 pathways for clusters 1 and 10. Classes 7, 8, and 9 were highly concordant in
terms of which pathways had enrichment. To explore the specific pathways marking
each class, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Classes 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, class 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Classes 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure3** and **supplementary figure S4**). Cluster 2 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
"$\beta$-catenin degradation", "cellular response to hypoxia", "regulation of
cell cycle" and "apoptosis" among others.

@@

Since pathways share some common driver genes, we investigated whether
tumors in each class  have multiple mutations across signaling pathways, or
whether a small number of driver genes account for enrichment. The split between
clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
**supplementary figure S4**). Furthermore, cluster 1 is distinguished
from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
_TP53_ panel of **Supplementary Figure S4**).

Cluster 2 exhibited the highest pathway enrichment scores of the ten major
classes, although there is a significant number of tumors spatially distributed
between cluster groupings 7-9 and 2 on the UMAP projection that have pathway
characteristics of both. The top ten distinguishing pathways of class 2 tumors
were "cellular response to hypoxia," "regulation of mitotic cell cycle,"
"assembly of the pre-replicative complex," "metabolism of polyamines,"
"degradation of $\beta$-catenin by the destruction complex", "tumor necrosis
factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0)
binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways
associated with aspects of hedgehog signaling. To determine once again whether
the genes driving pathway enrichment were commonly shared among these pathways
we plotted and performed unsupervised clustering on tumors with an individual
mutation matrix using every gene in each of the ten pathways, grouped by shared
pathway membership (**figure 3A**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity contributes to all the
aforementioned pathways and the majority of tumors in these classes have a
single mutation in one of these genes. We annotated the tumors in this class
according to tissue of origin to determine whether proteasome mutations have a
tissue-of-origin (or cancer-specific) bias, and indeed we could observe some
clustering. These data could explain how tissue-specific gene mutation
signatures might result in general cancer phenotypes by converging at the
pathway level.

@@

The proteasome plays key roles in the transduction of signals from the p53, WNT
and NFkB pathways and it and proteasomal inhibitors have been approved for
clinical use in some patient populations [@proteasome_review_Jang]. To determine
whether this class of proteins is broadly sensitive to cancer-promoting
mutations, we tabulated each gene as prognostic or not in the human protein
expression atlas [@protein_atlas] and found 49 positive and 10 negative
associations ( _i.e._ higher expression associated with unfavorable outcomes is
"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 3B**).
Second we looked at survival in mutations (**figure 3C**). Only a handful of proteasome mutations were significantly associated with favorable or unfavorable outcomes, probably owing to lack of sufficient power.

Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
majority (x out of y) are missense _vs._ various other deleterious classes (see
**figure 3D**), suggesting that, similar to other oncogenic driver classes such
as KRAS and PIK3C, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.
